---
title: "Biostat 212a Homework 1"
subtitle: "Due Jan 23, 2024 @ 11:59PM"
author: "YOUR NAME and UID"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    cache: false
---

```{r}
library(ISLR2)
library(GGally) # ggpairs function
library(tidyverse)
library(pROC)
```

## Filling gaps in lecture notes (10pts)

Consider the regression model
$$
Y = f(X) + \epsilon,
$$
where $\operatorname{E}(\epsilon) = 0$. 

### Optimal regression function

Show that the choice
$$
f_{\text{opt}}(X) = \operatorname{E}(Y | X)
$$
minimizes the mean squared prediction error
$$
\operatorname{E}[Y - f(X)]^2,
$$
where the expectations averages over variations in both $X$ and $Y$. (Hint: condition on $X$.)

**Solution**

Conditioning on $X$
\begin{eqnarray*}
\operatorname{E}[Y - f(X)]^2 &=& \operatorname{E}_X \operatorname{E}_{Y \mid X} \{ [Y - f(X)]^2 \mid X \}.
\end{eqnarray*}
Now at $X=x$,
\begin{eqnarray*}
\operatorname{E}_{Y \mid X} [ (Y - c)^2 \mid X = x ] &=& \int (y - c)^2 \, dF_{Y\mid X=x}(y)
\end{eqnarray*}
is minimized by $c(x) = \operatorname{E}_{Y \mid X = x}(Y \mid X = x)$. Therefore $f_{\text{opt}}$ minimizes the mean squared prediction error.

### Bias-variance trade-off

Given an estimate $\hat f$ of $f$, show that the test error at a $x_0$ can be decomposed as
$$
\operatorname{E}[y_0 - \hat f(x_0)]^2 = \underbrace{\operatorname{Var}(\hat f(x_0)) + [\operatorname{Bias}(\hat f(x_0))]^2}_{\text{MSE of } \hat f(x_0) \text{ for estimating } f(x_0)} + \underbrace{\operatorname{Var}(\epsilon)}_{\text{irreducible}},
$$
where the expectation averages over the variability in $y_0$ and $\hat f$.

**Solution** 

\begin{eqnarray*}
& & \operatorname{E}_{y_0,\hat f}(y_0 - \hat f(x_0))^2 \\
&=& \operatorname{E}_{y_0,\hat f} (y_0 - f(x_0) + f(x_0) - \hat f(x_0))^2 \\
&=& \operatorname{E}_{y_0,\hat f} (y_0 - f(x_0))^2 + 2 \operatorname{E}_{y_0,\hat f} (y_0 - f(x_0))(f(x_0) - \hat f(x_0)) + \operatorname{E}_{y_0,\hat f} (f(x_0) - \hat f(x_0))^2 \\
&=& \operatorname{Var}(\epsilon) + 0 + \operatorname{E}_{\hat f} (f(x_0) - \hat f(x_0))^2 \\
&=& \operatorname{Var}(\epsilon) + \operatorname{E}_{\hat f} (f(x_0) - \operatorname{E}_{\hat f} \hat f(x_0) + \operatorname{E}_{\hat f} \hat f(x_0) - \hat f(x_0))^2 \\
&=& \operatorname{Var}(\epsilon) + \operatorname{E}_{\hat f} (f(x_0) - \operatorname{E}_{\hat f} \hat f(x_0))^2 + 2 \operatorname{E}_{\hat f} [(f(x_0) - \operatorname{E}_{\hat f} \hat f(x_0))(\operatorname{E}_{\hat f} \hat f(x_0) - \hat f(x_0))] + \operatorname{E}_{\hat f} (\operatorname{E}_{\hat f} \hat f(x_0) - \hat f(x_0))^2 \\
&=& \operatorname{Var}(\epsilon) + (f(x_0) - \operatorname{E}_{\hat f} \hat f(x_0))^2 + 0 + \operatorname{Var}(\hat f(x_0)) \\
&=& \operatorname{Var}(\epsilon) + [\operatorname{Bias}(\hat f(x_0))]^2 + \operatorname{Var}(\hat f(x_0)).
\end{eqnarray*}

## ISL Exercise 2.4.3 (10pts)
We now revisit the bias-variance decomposition.

(a) Provide a sketch of typical (squared) bias, variance, training error, 
test error, and Bayes (or irreducible) error curves, on a single plot, 
as we go from less flexible statistical learning methods
towards more flexible approaches. The x-axis should represent
the amount of flexibility in the method, and the y-axis should
represent the values for each curve. There should be five curves.
Make sure to label each one.

**Solution**


(b) Explain why each of the five curves has the shape displayed in
part (a).

## ISL Exercise 2.4.4 (10pts)

You will now think of some real-life applications for statistical learning.

(a) Describe three real-life applications in which classification might
be useful. Describe the response, as well as the predictors. Is the
goal of each application inference or prediction? Explain your
answer.

**Solution** 

1. Among lung transplant patients, acute rejection is deadly. 
The goal of the study is to use each patients gene expression data from
blood to classify who will develop acute rejection who will not. The predictors
are gene expression, the outcome is the acute rejection. The primary goal is
prediction. 

2. Same problem as above, the goal is to identify genes that 
are differential expressed between reject and not reject. The primay goal is 
inference. 

(b) Describe three real-life applications in which regression might
be useful. Describe the response, as well as the predictors. Is the
goal of each application inference or prediction? Explain your
answer.

**Solution**

1. Identify the factors that are associated with kids' exam scores.
These factors include, class size, number of TAs, 
experiences of the teachers (years), kids' entrance exam score. Inference.

2. Patient Length of Stay (LOS) in hospital or ER.Prediction

(c) Describe three real-life applications in which cluster analysis
might be useful.

**Solution**

1. Identify subtypes of a complex diseases, e.g., diabetes. Features include
biomarkers, vital signs, medical history, etc. 

2. Identify ethnicity groups based on genetic data. 

3. Identify diseased areas in diagnostic images like X-rays that are abnormal. 

## ISL Exercise 2.4.10 (30pts)

Your can read in the `boston` data set directly from url <https://raw.githubusercontent.com/ucla-econ-425t/2023winter/master/slides/data/Boston.csv>. A documentation of the `boston` data set is [here](https://www.rdocumentation.org/packages/ISLR2/versions/1.3-2/topics/Boston).

::: {.panel-tabset}

#### Python

```{python}
import pandas as pd
import io
import requests

# Display all columns
pd.set_option('display.max_columns', None)

url = "https://raw.githubusercontent.com/ucla-econ-425t/2023winter/master/slides/data/Boston.csv"
s = requests.get(url).content
Boston = pd.read_csv(
  io.StringIO(s.decode('utf-8')), 
  index_col = 0, 
  # dtype = {'chas': 'category'}
  )
Boston
```

#### R

```{r}
library(tidyverse)

Boston <- read_csv("https://raw.githubusercontent.com/ucla-econ-425t/2023winter/master/slides/data/Boston.csv", col_select = -1) %>% 
  print(width = Inf)
```

:::

This exercise involves the Boston housing data set.


### (a) 

How many rows are in this data set? How many columns? What
do the rows and columns represent?

**Solution**

There are `r dim(Boston)[1]` rows and `r dim(Boston)[2]` columns in the `boston` data set. Each row is a suburb of Boston. Each column is a variable being measured on suburbs.

::: {.panel-tabset}

#### R

```{r}
library(gtsummary)

dim(Boston)
Boston %>% tbl_summary()
```

#### Python

```{python}
Boston.shape
```

:::


### (b) 

Make some pairwise scatterplots of the predictors (columns) in
this data set. Describe your findings.

**Solution**

We can use `pairplot` in Seaborn package for the pairwise scatter plots. It seems that `zn` is correlated with `rm` and `dis`. `indus` is correlated with `nox`.  

::: {.panel-tabset}


#### R

```{r}
library(GGally)

# Graphical summary
ggpairs(
  data = Boston,
  mapping = aes(alpha = 0.25),
  lower = list(continuous = "smooth")
  ) +
  labs(title = "Boston Data")
```

```{r}
# Numerical summary
summary(Boston)
```


#### Python

```{python}
import seaborn as sns
import matplotlib.pyplot as plt
# Set font sizes in plots
sns.set(font_scale = 2)

# Graphical summary
sns.pairplot(data = Boston)
```

```{python}
# Numerical summary
Boston.describe()
```

:::

### (c) 

Are any of the predictors associated with per capita crime rate?
If so, explain the relationship.

**Solution**

- `crim` seems to be highly correlated with `indus`, `nox`, `rad`, `tax`, and `lstat`, all with Pearson correlation larger than 0.4.

::: {.panel-tabset}



#### R

```{r}
cor(Boston, method = "pearson") %>% round(digits = 2)
```

#### Python

```{python}
round(Boston.corr(), 2)
```

:::

### (d) 

Do any of the census tracts of Boston appear to have particularly
high crime rates? Tax rates? Pupil-teacher ratios? Comment on
the range of each predictor.

**Solution**

- There seem to be some outliers in `crim` (>20). 

::: {.panel-tabset}

#### R

```{r}
# Numerical summary
Boston %>%
  select(crim, tax, ptratio) %>%
  summary()
```

#### Python

```{python}
cols = ['crim', 'tax', 'ptratio']
# Numerical summaries
Boston[cols].describe()
```

```{python}
plt.clf()
# Graphical summaries
for i, col in enumerate(cols):
    sns.histplot(data = Boston, x = col)
    plt.show()
    sns.boxplot(data = Boston, x = col)
    plt.show()
```



:::

### (e)

How many of the census tracts in this data set bound the Charles
river?

**Solution**

`r sum(Boston["chas"])` census tracts bound the Charles river.

::: {.panel-tabset}

#### R

```{r}
# Frequency table
table(Boston["chas"])
```


#### Python

```{python}
# Count the value 1
Boston['chas'].sum()
```

:::

### (f) 

What is the median pupil-teacher ratio among the towns in this
data set?

**Solution**

The median pupil-teacher ratio is `r median(Boston$ptratio)`.

::: {.panel-tabset}

#### R

```{r}
median(Boston$ptratio)
```

#### Python

```{python}
Boston['ptratio'].describe()
Boston['ptratio'].median()
```



:::

### (g)

Which census tract of Boston has lowest median value of owner-occupied homes? 
What are the values of the other predictors
for that census tract, and how do those values compare to the
overall ranges for those predictors? Comment on your findings.

**Solution**

There are two census tracts with the minimum `medv` `r min(Boston$medv)`. 
They both have relative high `ptratio`

::: {.panel-tabset}

#### R

```{r}
Boston %>%
  slice_min(medv) %>%
  print(width = Inf)
```

#### Python

```{python}
Boston[Boston['medv'] == Boston['medv'].min()]
```

:::


### (h) 

In this data set, how many of the census tracts average more than
seven rooms per dwelling? More than eight rooms per dwelling?
Comment on the census tracts that average more than eight rooms per dwelling.

**Solution**

There are `r Boston %>% filter(rm > 8) %>% count()` census tracts with more than 8 rooms per dwelling. 

::: {.panel-tabset}

#### R

```{r}
Boston %>%
  filter(rm > 8) %>%
  print(width = Inf)
```

#### Python

```{python}
Boston[Boston['rm'] > 7].shape[0]
Boston[Boston['rm'] > 8].shape[0]
```

```{python}
plt.clf()
sns.histplot(data = Boston, x = 'rm')
plt.show()
```

```{python}
Boston[Boston['rm'] > 8]
```

:::

## ISL Exercise 3.7.3 (12pts)

Suppose we have a data set with five predictors, 
$X_1 = \text{GPA}$, 
$X_2 = \text{IQ}$, 
$X_3 = \text{Level (1 for College and 0 for High School)}$, 
$X_4 = \text{Interaction between GPA and IQ}$, and 
$X_5 = \text{Interaction between GPA and Level}$. 
The response is starting salary after graduation (in thousands of dollars). 
Suppose we use least squares to fit the model, and get 
$\hat\beta_0 = 50$, 
$\hat\beta_1 = 20$,
$\hat\beta_2 = 0.07$, 
$\hat\beta_3 = 35$,
$\hat\beta_4 = 0.01$, 
$\hat\beta_5 = −10$.

The model is
$$
\text{salary} \approx \beta_0 + \beta_1 \times \text{GPA} + \beta_2 \times \text{IQ} + \beta_3 \times \text{Level} + \beta_4 \times (\text{GPA} \times \text{IQ}) + \beta_5 \times (\text{GPA} \times \text{Level})
$$




For a fixed value of IQ and GPA, collage students earn
$$
\beta_3 + \beta_5 * (\text{GPA}) = 35 - 10 \times \text{GPA}
$$
more than high school graduates. Depending on \text{GPA}, colleges graduates can earn more or less than high school graduates. If GPA is high enough, then high school graduates earn more.


### (a) 

Which answer is correct, and why?

- For a fixed value of IQ and GPA, high school graduates earn
more, on average, than college graduates. 

- For a fixed value of IQ and GPA, college graduates earn
more, on average, than high school graduates.

- For a fixed value of IQ and GPA, high school graduates earn
more, on average, than college graduates provided that the
GPA is high enough.

- For a fixed value of IQ and GPA, college graduates earn
more, on average, than high school graduates provided that
the GPA is high enough.

### (b)

Predict the salary of a college graduate with IQ of 110 and a GPA of 4.0.

**Solution**

For a college student with IA of 110 and a GPA of 4.0, 
$$
\hat y = \hat \beta_0 + \hat \beta_1 \times 4 + \hat \beta_2 \times 110 + \hat \beta_3 + \hat \beta_4 \times (4 \times 110) + \hat \beta_5 \times (4) = ???
$$

### (c) 

True or false: Since the coeﬀicient for the GPA/IQ interaction
term is very small, there is very little evidence of an interaction
effect. Justify your answer.

**Solution**

Depending on the standard error, the p-value of interaction term can be very small thus significant.


## ISL Exercise 3.7.15 (20pts)


This problem involves the `Boston` data set, which we saw in the lab
for this chapter. We will now try to predict per capita crime rate
using the other variables in this data set. In other words, per capita
crime rate is the response, and the other variables are the predictors.


### (a)

For each predictor, fit a simple linear regression model to predict
the response. Describe your results. In which of the models is
there a statistically significant association between the predictor
and the response? Create some plots to back up your assertions.

**Solution**

- Retrieve simple linear regression (SLR) coefficients by loop

::: {.panel-tabset}

#### R


```{r}
library(gtsummary)
crim.fit <- lm(crim ~ . , data = Boston)
summary(crim.fit)
```


#### Python

```{python}
import statsmodels.api as sm
import statsmodels.formula.api as smf

# Fit separate simple linear regressions
cols = Boston.columns.difference(['crim']) # all cols except crim
cols
slr_coefs = [
  smf.ols(formula = 'crim ~ ' + col, data = Boston).fit().params[1]
  for col in cols
  ]
slr_coefs  
res_df = pd.DataFrame({'slr_betahat': slr_coefs}, index = cols)
res_df
```

:::


### (b) 

Fit a multiple regression model to predict the response using all of the predictors. 
Describe your results. For which predictors can we reject the null hypothesis 
$\text{H}_0: \beta_j = 0$?

**Solution**

Retrieve multiple linear regression (MLR) coefficients and contrast to SLR coefficients. 

::: {.panel-tabset}
#### R

```{r}
lm(crim ~ ., data = Boston) %>%
  tbl_regression() %>%
  bold_labels() %>%
  bold_p(t = 0.05)
```

#### Python

```{python}
mlr_coefs = smf.ols(formula = 'crim ~ ' + "+".join(cols), data = Boston).fit().params.to_frame(name = 'mlr_betahat')
res_df = res_df.join(mlr_coefs) # left-join by default
res_df
```

```{python}
sns.relplot(
  data = res_df, 
  x = 'slr_betahat', 
  y = 'mlr_betahat', 
  height = 8).set(
  xlabel = r'SLR $\hat{\beta}_j$',
  ylabel = r'MLR $\hat{\beta}_j$',
)
```
:::

### (c)

How do your results from (a) compare to your results from (b)?
Create a plot displaying the univariate regression coeﬀicients
from (a) on the x-axis, and the multiple regression coeﬀicients
from (b) on the y-axis. That is, each predictor is displayed as a
single point in the plot. 
Its coeﬀicient in a simple linear regression model is shown on the x-axis, 
and its coeﬀicient estimate in the multiple linear regression model 
is shown on the y-axis.


### (d) 

Is there evidence of non-linear association between any of the
predictors and the response? To answer this question, for each
predictor X, fit a model of the form

$$
Y= \beta_0 + \beta_1X_1 + \beta_2X_2 + \beta_3X_3 + \epsilon.
$$

## Bonus question (20pts)

For multiple linear regression, show that $R^2$ is equal to the correlation between the response vector $\mathbf{y} = (y_1, \ldots, y_n)^T$ and the fitted values $\hat{\mathbf{y}} = (\hat y_1, \ldots, \hat y_n)^T$. That is
$$
R^2 = 1 - \frac{\text{RSS}}{\text{TSS}} = [\operatorname{Cor}(\mathbf{y}, \hat{\mathbf{y}})]^2.
$$

### Solution



- Recall that 
\begin{eqnarray*}
 \operatorname{Cor} (\mathbf{x}, \mathbf{y}) & = & \frac{\sum_i (x_i - \bar{x}) (y_i - \bar{y})} {\sqrt{\sum_i(x_i - \bar{x})^2 \sum_i(y_i - \bar{y})^2}}\\
 & = & \frac{( \mathbf{x} - \frac{\mathbf{J}}{n}\mathbf{x} )^T (\mathbf{y} - \frac{\mathbf{J}}{n}\mathbf{y})}{\sqrt{(\mathbf{x} - \frac{\mathbf{J}}{n}\mathbf{x})^T (\mathbf{x} - \frac{\mathbf{J}}{n}\mathbf{x}) (\mathbf{y} - \frac{\mathbf{J}}{n}\mathbf{y})^T(\mathbf{y} - \frac{\mathbf{J}}{n}\mathbf{y})}}\\
\end{eqnarray*}
where 
$$
\mathbf{J} = \mathbf{1}\mathbf{1}^T = \begin{pmatrix} 1 & 1 & \cdots & 1 \\ 1 & 1 & \cdots & 1 \\ \vdots & \vdots & \ddots & \vdots \\ 1 & 1 & \cdots & 1 \end{pmatrix}_{n \times n}.
$$

- And for $\operatorname{R}^2$
$$
 \operatorname{R}^2  = 1-\frac{\text{RSS}} {\text{TSS}} = 1 - \frac{\sum_i (y_i- \hat{y}_i)^2}{\sum_i (y_i- \bar{y}_i)^2}
$$
and 
$$
 \operatorname{R}^2  = 1 - \frac{\sum_i (y_i- \hat{y}_i)^2}{\sum_i (y_i- \bar{y}_i)^2} = 1 - \frac{(\mathbf{Y}-\mathbf{H}\mathbf{Y})^T(\mathbf{Y}-\mathbf{H}\mathbf{Y})}{(\mathbf{Y}-\frac{\mathbf{J}}{n}\mathbf{Y})^T(\mathbf{Y}-\frac{\mathbf{J}}{n}\mathbf{Y})} = 1 - \frac{\mathbf{Y}^T(\mathbf{I}-\mathbf{H})^T(\mathbf{I}-\mathbf{H})\mathbf{Y}}{\mathbf{Y}^T(\mathbf{I}-\frac{\mathbf{J}}{n})^T(\mathbf{I}-\frac{\mathbf{J}}{n})\mathbf{Y}}
$$
Therefore,
$$
 \operatorname{R}^2  = 1 - \frac{\mathbf{Y}^T(\mathbf{I}-\mathbf{H})\mathbf{Y}}{\mathbf{Y}^T(\mathbf{I}-\frac{\mathbf{J}}{n})\mathbf{Y}}
$$
where $\mathbf{H} = \mathbf{X}(\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T$ is a projection matrix and so as $\mathbf{I}-\mathbf{H}$.

Now, I will use $H_{\mathbf{1}_n}$ to be $\frac{\mathbf{J}}{n}$.

\begin{eqnarray*}
[\operatorname{Cor}(\mathbf{y}, \hat{\mathbf{y}})]^2 &=& \frac{[\langle (I - H_{\mathbf{1}_n}) H_{\mathbf{X}} \mathbf{y}, (I - H_{\mathbf{1}_n}) \mathbf{y} \rangle]^2}{\|(I - H_{\mathbf{1}_n}) H_{\mathbf{X}} \mathbf{y}\|^2 \|(I - H_{\mathbf{1}_n}) \mathbf{y}\|^2} \\
&=& \frac{[\mathbf{y}^T H_{\mathbf{X}} (I - H_{\mathbf{1}_n}) \mathbf{y}]^2}{[\mathbf{y}^T H_{\mathbf{X}} (I - H_{\mathbf{1}_n}) H_{\mathbf{X}} \mathbf{y}][\mathbf{y}^T (I - H_{\mathbf{1}_n}) \mathbf{y}]} \\
& & (\text{now } \mathbf{y}^T H_{\mathbf{X}} (I - H_{\mathbf{1}_n}) H_{\mathbf{X}} \mathbf{y} = \mathbf{y}^T H_{\mathbf{X}} \mathbf{y} - \mathbf{y}^T H_{\mathbf{1}_n} H_{\mathbf{X}} \mathbf{y} = \mathbf{y}^T H_{\mathbf{X}} (I - H_{\mathbf{1}_n}) \mathbf{y}) \\
&=& \frac{|\mathbf{y}^T H_{\mathbf{X}} (I - H_{\mathbf{1}_n}) \mathbf{y}|}{\mathbf{y}^T (I - H_{\mathbf{1}_n}) \mathbf{y}} \\
&=& \frac{|\mathbf{y}^T H_{\mathbf{X}} \mathbf{y} - \mathbf{y}^T H_{\mathbf{1}_n} \mathbf{y}|}{\mathbf{y}^T (I - H_{\mathbf{1}_n}) \mathbf{y}} \\
&=& \frac{\mathbf{y}^T (\mathbf{I} - H_{\mathbf{1}_n}) \mathbf{y} - \mathbf{y}^T (\mathbf{I} - H_{\mathbf{X}}) \mathbf{y}}{\mathbf{y}^T (I - H_{\mathbf{1}_n}) \mathbf{y}} \\
&=& 1 - \frac{\text{RSS}}{\text{TSS}}.
\end{eqnarray*}


<!--
- Now we have
$$
\operatorname{Cor}(\mathbf{Y}, \hat{\mathbf{Y}})^2  =  \frac{[( \mathbf{Y} - \frac{\mathbf{J}}{n}\mathbf{Y} )^T (\hat{\mathbf{Y}} - \frac{\mathbf{J}}{n}\hat{\mathbf{Y}})]^2}{[(\mathbf{Y} - \frac{\mathbf{J}}{n}\mathbf{Y})^T (\mathbf{Y} - \frac{\mathbf{J}}{n}\mathbf{Y})][ (\hat{\mathbf{Y}} - \frac{\mathbf{J}}{n}\hat{\mathbf{Y}})^T (\hat{\mathbf{Y}} - \frac{\mathbf{J}}{n}\hat{\mathbf{Y})]}}  
$$     

where 
$$
\hat{\mathbf{Y}} = \mathbf{X}\hat{\boldsymbol{\beta}} = \mathbf{X}(\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{Y} = \mathbf{H}\mathbf{Y}
$$
-->

## TA Reports 
### James
- Q1) Many students had a correct solution, a few assumed the defined optimal solution was optimal and used that to prove that it was optimal which is circular. 

- Q2) Students did generally well here. A few did not have their graph display, one had test and train error inverted. 

- Q3) Almost all students got this entirely correct a few did not answer it in full. 

- Q4) 4d) was the major issue with many not identifying or justifying "exceptional values." In 4g several only identified a single minimum when two tracts were tied. 

- Q5) The wording tripped up some as they selected IV instead of III). One student got the calculation wrong for 5.2. Several got 5.3 wrong as they thought a small coefficient was insignificant. 

- Q6) In part c) many students provided a graph without explanation of how the coefficients compared.

- Bonus) Many students asserted that $\text{cov}(y, \hat y) = \text{var}(\hat y)$ without justification. For a question of this type I figured they needed to actually prove this under the assumptions of multiple linear regression. Only two students gave a proof I accepted as sufficiently complete. 

### Yingxin

- Q1: For Q1.1, some students were unclear on how to begin their proof. Additionally, a few students were unable to complete the proof for Q1.2.

- Q2: Most students were able to create the sketch using either R or handwriting, but some encountered issues displaying their plots in the HTML file due to incorrect settings.

- Q3: All students provided some applications.

- Q4: Most students completed this question well. However, many were unfamiliar with the concept of “outliers” in Q4.4, and some only identified one of the two lowest medv tracts in Q4.7.

- Q5: While most students performed well, some selected extra choices instead of the correct option (iii) in Q5.1, and two students miscalculated their results for Q5.2.

- Q6: All students completed this question correctly.

- Bonus: All students who attempted the bonus missed the proof of $\sum {(y_i - \hat y_i)(\hat y_i - \bar y)} = 0$ or a similar key step.



