---
title: "Domain Shift, Transfer Learning, and Model Recalibration"
subtitle: "Biostat 274"
author: "Dr. Jin Zhou @ UCLA"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    message: FALSE
    cache: false
---

Credit: This note heavily uses material from

- Huang et al. (2020). [_A tutorial on calibration measurements and calibration models for clinical prediction models_](https://doi.org/10.1093/jamia/ocz228). *Journal of the American Medical Informatics Association*, 27(4), 621–633.

- Van Calster B, et al. (2016). A calibration hierarchy for risk models was defined: from utopia to empirical data. *J Clin Epidemiol*, 74, 167–176.

- Kouw & Loog (2018). [_An Introduction to Domain Adaptation and Transfer Learning_](https://arxiv.org/abs/1812.11806).

- Chip Huyen. [_Data Distribution Shifts and Monitoring_](https://huyenchip.com/2022/02/07/data-distribution-shifts-and-monitoring.html).

---

## Motivation

### The Core Problem

- Most clinicians can recall seeing a patient who was listed as 50 years old but appeared decades older due to chronic illness — "appearing older than stated age."

- Experienced clinicians can **calibrate their mental models** to account for such discrepancies. Predictive models cannot do this without further instruction.

- When predictive models are built on one population and applied to another, blind application can result in large **residuals** (large differences between model estimates and true outcomes) — even when average performance looks acceptable.

### A Motivating Example: Dementia Risk Prediction

- One study evaluated calibration of several models for predicting risk of dementia in community-dwelling populations.

- At a **predicted risk of 40%**, the observed incidence was still only **10%** — a 30 percentage point overestimation.

- For an individual patient, this magnitude of error could lead to dramatically different clinical decisions.

### Another Example: Cardiovascular Risk

- The American College of Cardiology / American Heart Association recommend prescribing statins for patients with cardiovascular risk **> 7.5%** and considering treatment for those between **5–7.5%**.

- A miscalibrated model assigning 8% risk to a patient whose true risk is 4% could lead to unnecessary statin prescription — and vice versa.

> **Key insight:** Even when a model's *average* predicted risk equals the observed prevalence, there could be severe miscalibration in specific subgroups. Accurate, well-calibrated estimates are necessary to allocate resources appropriately.

### Discrimination vs. Calibration: The Critical Distinction

- **Discrimination** refers to the ability of a model to *rank* patients according to risk — measured by the AUROC (area under the ROC curve), also known as the concordance index.

- **Calibration** refers to the *agreement* between predicted probabilities and observed outcomes — it captures whether predicted risks are numerically accurate.

- **These are independent properties.** Three models with identical AUROC can have dramatically different calibration. See Figure 1 from Huang et al. (2020):

    - Model A: original estimates — well-calibrated.
    - Model B: original estimates divided by 10 — perfectly ordered but grossly underestimates.
    - Model C: recalibrated estimates — improved calibration.
    - **All three have identical AUROC = 0.785.**

- One systematic review found that while **63% of published models** included a discrimination measure, only **36%** included a calibration measure.

> **Implication:** Reporting only AUROC is insufficient for clinical decision making. A model may rank patients correctly but assign completely wrong absolute risk values.

### The Calibration Hierarchy (Van Calster et al., 2016)

Calibration can be evaluated at multiple levels, from coarse to fine:

1. **Mean calibration (calibration-in-the-large):** Does the average predicted probability equal the observed event rate?

2. **Weak calibration:** Does the calibration slope equal 1 and intercept equal 0?

3. **Moderate calibration:** Within subgroups defined by predicted risk, do predicted and observed rates agree?

4. **Strong calibration:** For every individual, does the predicted probability equal the true probability?

Strong calibration is an ideal that is rarely achievable in practice. Most clinical work focuses on levels 1–3.

---

## Types of Calibration Metrics

There is no single best method to measure calibration. Each method has specific strengths and limitations. Below we present the most commonly used metrics.

### Brier Score

The **Brier score** is the mean squared error between predicted probabilities and actual binary outcomes:

$$
\text{Brier Score} = \frac{1}{N} \sum_{i=1}^{N} (E_i - O_i)^2
$$

where $E_i$ is the predicted probability for patient $i$ and $O_i \in \{0, 1\}$ is the actual outcome.

- A Brier score of **0** indicates perfect prediction; a score of **0.25** is the score of a model that always predicts 0.5.

- The Brier score is a **combined measure** of discrimination and calibration — a lower score does not necessarily mean better calibration alone.

- It can be **decomposed** into a calibration component and a refinement (discrimination) component — the calibration component can be extracted via the Spiegelhalter z-test.

```r
library(DescTools)
BrierScore(predicted_probs, actual_outcomes)
```

### Spiegelhalter's Z-Test

An extension of the Brier score that isolates calibration. The test statistic is:

$$
Z(E, O) = \frac{\sum_{i=1}^{N}(O_i - E_i)(1 - 2E_i)}{\sqrt{\sum_{i=1}^{N}(1-2E_i)^2 E_i(1-E_i)}}
$$

- If $|Z| > z_{1-\alpha/2}$ (e.g., $z_{0.975} = 1.96$), the model is considered miscalibrated at level $\alpha = 0.05$.

- Provides a formal statistical test for calibration — useful as a hypothesis test alongside visualizations.

```r
library(rms)
val.prob(predicted_probs, actual_outcomes)  # includes Spiegelhalter z
```

### Hosmer-Lemeshow (H-L) Test

The most widely used proxy for calibration in biomedical literature (874+ PubMed mentions in the past 5 years).

**Two variants:**

- **H-L C-statistic:** Divide patients into $g$ groups (typically deciles) with **equal numbers of patients** per group. Compare expected vs. observed events within each group.

- **H-L H-statistic:** Divide patients into groups based on **equal-width intervals** of predicted probability (e.g., [0, 0.1), [0.1, 0.2), ...).

The test statistic follows a $\chi^2$ distribution with $g-2$ degrees of freedom:

$$
\text{H-L} = \sum_{i=1}^{g} \left[ \frac{(O_{s,i} - E_{s,i})^2}{E_{s,i}} + \frac{(O_{f,i} - E_{f,i})^2}{E_{f,i}} \right]
$$

- $p \geq 0.10$: adequate calibration
- $0.05 < p < 0.10$: borderline
- $p < 0.05$: significant miscalibration

```{r, eval = F}
library(ResourceSelection)
hoslem.test(actual_outcomes, predicted_probs, g = 10)
```

**Limitations:**

- Susceptible to increasing power as sample size increases — may reject calibration in large datasets even for trivial deviations.
- H-L C and H-L H can give contradictory results.
- Choice of number of bins is arbitrary and can alter the p-value.

### Reliability Diagram (Calibration Curve)

A visualization of calibration that uses the same groupings as the H-L test. For each bin, the **mean predicted probability** is plotted against the **observed event rate**. A perfectly calibrated model produces a 45-degree diagonal line.

- The reliability diagram reveals the **direction** of miscalibration (underestimation, overestimation) and **where** it occurs.

- Deviations above the diagonal indicate underestimation of risk; deviations below indicate overestimation.

```{r, eval = F}
library(PresenceAbsence)
calibration.plot(actual_outcomes, predicted_probs, N.bins = 10)
```

**Limitation:** Not a continuous curve — no information can be derived between the plotted points. Can be hard to read when estimates are clustered in a narrow range.

### Expected Calibration Error (ECE) and Maximum Calibration Error (MCE)

Binning-based scalar summaries of miscalibration, popular in the machine learning literature:

$$
\text{ECE} = \sum_{k=1}^{K} P(k) \cdot |o_k - e_k|, \quad \text{MCE} = \max_{k=1,\ldots,K} |o_k - e_k|
$$

where $P(k)$ is the fraction of patients in bin $k$, $o_k$ is the observed event rate, and $e_k$ is the mean predicted probability.

- **ECE** measures the *average* calibration error, weighted by bin size.
- **MCE** measures the *worst-case* calibration error across bins.
- Both are intuitive and easy to compute.

**Limitation:** No statistical test to determine if a given ECE/MCE value is "too large" — requires context-dependent thresholds.

### Cox Intercept and Slope

A regression-based approach that avoids binning entirely. The binary outcome is regressed on the log-odds of the predicted probability:

$$
\text{logit}\{P(O=1)\} = \alpha + \beta \cdot \text{logit}(E)
$$

- **Slope** $\beta = 1$: perfect calibration.
- $\beta > 1$: underestimation of high risk, overestimation of low risk (predictions too extreme).
- $\beta < 1$: overestimation of high risk, underestimation of low risk (predictions too conservative).
- **Intercept** $\alpha = 0$: good overall calibration.
- $\alpha > 0$: systematic underestimation (predicted probabilities are too low on average).
- $\alpha < 0$: systematic overestimation.

```{r, eval = F}
cal_model <- glm(actual ~ offset(logit_predicted), 
                 family = binomial, data = df)  # intercept-only recalibration

cal_model2 <- glm(actual ~ logit_predicted, 
                  family = binomial, data = df)  # slope + intercept
```

**Limitation:** Cox slope and intercept can both be 1 and 0 respectively while deviations in specific regions still exist — if miscalibrations "cancel out" in the linear regression.

### Integrated Calibration Index (ICI)

The ICI extends Cox's approach by using a **locally weighted regression smoother (Loess)** instead of a linear model:

$$
\text{ICI} = \frac{1}{N} \sum_{i=1}^{N} |E_i - \hat{p}(E_i)|
$$

where $\hat{p}(E_i)$ is the Loess-smoothed estimate of the true probability at $E_i$.

- Can detect miscalibration in specific regions that Cox slope/intercept misses (e.g., when overestimation in one range is cancelled by underestimation in another).

- Well-calibrated models have ICI close to 0.

```{r, eval= F}
library(rms)
val.prob(predicted_probs, actual_outcomes)  # includes ICI
```

**Limitation:** Requires choice of Loess bandwidth parameter. Less intuitive than binning methods for clinical audiences.

---

## Summary: Pros and Cons of Calibration Metrics

| Metric | Pros | Cons |
|:--|:--|:--|
| **Brier score** | Easy to compute; widely reported | Combines discrimination + calibration; components not easily separable |
| **Spiegelhalter z-test** | Formal statistical test; isolates calibration from Brier | Not intuitive; less commonly understood |
| **Hosmer-Lemeshow test** | Most widely used in biomedicine; p-value guides interpretation | Power increases with $n$; bin count is arbitrary; C vs H can disagree |
| **Reliability diagram** | Visualizes direction and location of miscalibration | Not a continuous curve; hard to read with clustered predictions |
| **ECE / MCE** | Intuitive scalar summaries; popular in ML | No formal statistical test; threshold is arbitrary |
| **Cox slope & intercept** | No binning needed; interprets direction of miscalibration | Misses regional miscalibration that cancels out linearly |
| **ICI** | Captures regional miscalibration missed by Cox | Requires Loess bandwidth choice; less intuitive |

> **Practical recommendation:** Use at least one visual (reliability diagram) and one scalar with a statistical test (Spiegelhalter z or H-L). Report Cox intercept and slope to characterize direction of any miscalibration. Use ICI for a more sensitive check.

---

## Figure 1 Reproduction: Same AUROC, Different Calibration

This reproduces the core insight of Huang et al. (2020) Figure 1: **three models can have identical AUROC but dramatically different calibration.**

We simulate data following the paper's setup (5,000 samples, 15% event rate, logistic regression and a miscalibrated SVM-like model).

```{r}
#| label: fig-auroc-vs-calibration
#| fig-cap: "Illustration that AUROC is insensitive to calibration. Three models (original, divided-by-10, recalibrated) have identical AUROC = 0.785 but very different predicted probability distributions."
#| code-fold: true

library(tidyverse)
library(tidymodels)
library(ResourceSelection)
library(patchwork)

set.seed(274)

# ── Simulate data (mirroring Huang et al. 2020) ──────────────────────────────
n       <- 2000
x1      <- rnorm(n); x2 <- rnorm(n); x3 <- rbinom(n, 1, 0.4)
log_odds <- -2 + 0.8*x1 + 0.6*x2 + 0.5*x3
true_prob <- plogis(log_odds)
y         <- rbinom(n, 1, true_prob)

# Split: 60% train, 20% calibration, 20% test
idx       <- sample(1:n)
train_idx <- idx[1:1200]
cal_idx   <- idx[1201:1600]
test_idx  <- idx[1601:2000]

df       <- tibble(y, x1, x2, x3)
df_train <- df[train_idx, ]
df_cal   <- df[cal_idx, ]
df_test  <- df[test_idx, ]

# ── Fit logistic regression (well-calibrated) ────────────────────────────────
lr_fit   <- glm(y ~ x1 + x2 + x3, data = df_train, family = binomial)
p_lr     <- predict(lr_fit, newdata = df_test, type = "response")

# ── Model A: original LR predictions ────────────────────────────────────────
p_A <- p_lr

# ── Model B: predictions divided by 10 (grossly miscalibrated) ──────────────
p_B <- p_lr / 10

# ── Model C: "recalibrated" (push toward 0/1 extremes using logit scaling) ──
p_C <- plogis(qlogis(p_lr) * 1.5)

# ── AUROC for all three ──────────────────────────────────────────────────────
y_test <- df_test$y

auroc <- function(pred, truth) {
  roc_auc(tibble(truth = factor(truth, levels = c("1","0")), prob = pred),
          truth, prob)$.estimate
}

cat(sprintf("AUROC Model A (original):     %.3f\n", auroc(p_A, y_test)))
cat(sprintf("AUROC Model B (divided by 10):%.3f\n", auroc(p_B, y_test)))
cat(sprintf("AUROC Model C (rescaled):     %.3f\n", auroc(p_C, y_test)))

# ── ROC curves ───────────────────────────────────────────────────────────────
roc_df <- bind_rows(
  roc_curve(tibble(truth=factor(y_test,levels=c("1","0")),prob=p_A), truth, prob) %>% mutate(model="A: Original"),
  roc_curve(tibble(truth=factor(y_test,levels=c("1","0")),prob=p_B), truth, prob) %>% mutate(model="B: Divided by 10"),
  roc_curve(tibble(truth=factor(y_test,levels=c("1","0")),prob=p_C), truth, prob) %>% mutate(model="C: Rescaled")
)

p_roc <- ggplot(roc_df, aes(x = 1-specificity, y = sensitivity, color = model, linetype = model)) +
  geom_line(linewidth = 0.9) +
  geom_abline(linetype = "dotted", color = "gray60") +
  scale_color_manual(values = c("#2196F3","#F44336","#4CAF50")) +
  labs(title = "ROC Curves: All three models overlap",
       subtitle = sprintf("All AUROCs ≈ %.3f", auroc(p_A, y_test)),
       x = "1 - Specificity", y = "Sensitivity", color = NULL, linetype = NULL) +
  theme_bw() + theme(legend.position = "bottom")

# ── Calibration curves ───────────────────────────────────────────────────────
make_cal_df <- function(pred, truth, label) {
  tibble(pred = pred, truth = truth) %>%
    mutate(bin = cut(pred, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>%
    group_by(bin) %>%
    summarise(mean_pred = mean(pred), mean_obs = mean(truth), n = n(), .groups = "drop") %>%
    mutate(model = label)
}

cal_all <- bind_rows(
  make_cal_df(p_A, y_test, "A: Original"),
  make_cal_df(p_B, y_test, "B: Divided by 10"),
  make_cal_df(p_C, y_test, "C: Rescaled")
)

p_cal <- ggplot(cal_all, aes(x = mean_pred, y = mean_obs, color = model)) +
  geom_line(linewidth = 0.9) +
  geom_point(aes(size = n)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  scale_color_manual(values = c("#2196F3","#F44336","#4CAF50")) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "Calibration Curves: Dramatically different",
       subtitle = "Diagonal = perfect calibration",
       x = "Mean Predicted Probability", y = "Observed Event Rate",
       color = NULL, size = "N in bin") +
  theme_bw() + theme(legend.position = "bottom")

p_roc + p_cal
```

**Key takeaway:** Model B (divided by 10) stacks perfectly on top of Model A in the ROC plot, but its calibration curve is completely flat near zero — it never assigns probabilities above 0.05 regardless of how high the true risk is.

---

## Figure 2: All Calibration Metrics on Simulated Data

This demonstrates all seven calibration metrics on a well-calibrated model (LR) versus a miscalibrated model.

### Setup: Well-calibrated vs. Miscalibrated Models

```{r}
#| label: setup-models
#| code-fold: true

# ── Well-calibrated: LR ──────────────────────────────────────────────────────
p_well <- p_A   # original LR predictions — well calibrated

# ── Miscalibrated: compress predictions toward 0 (like an SVM) ──────────────
# This mimics the SVM behavior in the paper: underestimates high risks
p_misc <- plogis(qlogis(p_A) * 0.3 - 1.5)  # compressed + shifted down

cat("=== Well-calibrated model ===\n")
cat(sprintf("  Mean predicted: %.3f  |  Observed rate: %.3f\n",
            mean(p_well), mean(y_test)))

cat("\n=== Miscalibrated model ===\n")
cat(sprintf("  Mean predicted: %.3f  |  Observed rate: %.3f\n",
            mean(p_misc), mean(y_test)))
```

### Metric 1: Brier Score

```{r}
#| label: brier-score

brier_well <- mean((p_well - y_test)^2)
brier_misc <- mean((p_misc - y_test)^2)

cat(sprintf("Brier score — Well-calibrated: %.4f\n", brier_well))
cat(sprintf("Brier score — Miscalibrated:   %.4f\n", brier_misc))
cat(sprintf("Null model (always predict mean): %.4f\n",
            mean((mean(y_test) - y_test)^2)))
```

### Metric 2: Hosmer-Lemeshow Test

```{r}
#| label: hl-test

library(ResourceSelection)

hl_well <- hoslem.test(y_test, p_well, g = 10)
hl_misc <- hoslem.test(y_test, p_misc, g = 10)

cat("=== Hosmer-Lemeshow Test ===\n")
cat(sprintf("Well-calibrated: X² = %.2f, p = %.4f  %s\n",
            hl_well$statistic, hl_well$p.value,
            ifelse(hl_well$p.value < 0.05, "⚠ MISCALIBRATED", "✓ OK")))
cat(sprintf("Miscalibrated:   X² = %.2f, p = %.4f  %s\n",
            hl_misc$statistic, hl_misc$p.value,
            ifelse(hl_misc$p.value < 0.05, "⚠ MISCALIBRATED", "✓ OK")))
```

### Metric 3: Cox Intercept and Slope

```{r}
#| label: cox-intercept-slope

cox_stats <- function(pred, truth, label) {
  df_cox    <- tibble(y = truth, logit_p = qlogis(pmin(pmax(pred, 1e-6), 1-1e-6)))
  
  # intercept only (fix slope = 1) → calibration-in-the-large
  m_int  <- glm(y ~ offset(logit_p), data = df_cox, family = binomial)
  
  # slope + intercept → weak calibration
  m_both <- glm(y ~ logit_p, data = df_cox, family = binomial)
  
  cat(sprintf("\n--- %s ---\n", label))
  cat(sprintf("  Intercept (α): %.3f  [ideal = 0]\n",
              coef(m_int)[["(Intercept)"]] ))
  cat(sprintf("  Slope     (β): %.3f  [ideal = 1]\n",
              coef(m_both)[["logit_p"]]))
  cat(sprintf("  Interpretation: %s\n",
              case_when(
                coef(m_both)[["logit_p"]] > 1.1 ~ "Underestimates high risk, overestimates low risk",
                coef(m_both)[["logit_p"]] < 0.9 ~ "Overestimates high risk, underestimates low risk",
                TRUE ~ "Slope near 1 — good weak calibration"
              )))
}

cox_stats(p_well, y_test, "Well-calibrated model")
cox_stats(p_misc, y_test, "Miscalibrated model")
```

### Metric 4: Reliability Diagram (Calibration Curve)

::: {#fig-reliability}

```{r}
#| label: fig-reliability-diagram
#| fig-cap: "Reliability diagrams for well-calibrated (blue) and miscalibrated (red) models. The miscalibrated model systematically underestimates risk — its curve lies above the diagonal, consistent with Figure 3 in Huang et al. (2020)."

cal_well <- make_cal_df(p_well, y_test, "Well-calibrated (LR)")
cal_misc <- make_cal_df(p_misc, y_test, "Miscalibrated")
cal_both <- bind_rows(cal_well, cal_misc)

ggplot(cal_both, aes(x = mean_pred, y = mean_obs, color = model)) +
  geom_line(linewidth = 1.0) +
  geom_point(aes(size = n)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40", linewidth = 0.8) +
  annotate("text", x = 0.7, y = 0.65, label = "Perfect calibration",
           color = "gray40", size = 3.2, angle = 35) +
  scale_color_manual(values = c("#2196F3", "#F44336")) +
  scale_x_continuous(limits = c(0, 0.8)) +
  scale_y_continuous(limits = c(0, 0.8)) +
  labs(title = "Reliability Diagram",
       subtitle = "Miscalibrated model: curve lies above diagonal → underestimates risk",
       x = "Mean Predicted Probability", y = "Observed Event Rate",
       color = NULL, size = "N in bin") +
  theme_bw() +
  theme(legend.position = "bottom")
```

:::

### Metric 5: ECE and MCE

```{r}
#| label: ece-mce

compute_ece_mce <- function(pred, truth, K = 10) {
  bins <- cut(pred, breaks = seq(0, 1, length.out = K+1), include.lowest = TRUE)
  df   <- tibble(pred, truth, bins) %>%
    group_by(bins) %>%
    summarise(mean_pred = mean(pred), mean_obs = mean(truth),
              n = n(), .groups = "drop") %>%
    mutate(w = n / sum(n), err = abs(mean_pred - mean_obs))
  
  list(ECE = sum(df$w * df$err, na.rm = TRUE),
       MCE = max(df$err, na.rm = TRUE))
}

ece_well <- compute_ece_mce(p_well, y_test)
ece_misc <- compute_ece_mce(p_misc, y_test)

cat("=== ECE and MCE ===\n")
cat(sprintf("Well-calibrated: ECE = %.4f | MCE = %.4f\n",
            ece_well$ECE, ece_well$MCE))
cat(sprintf("Miscalibrated:   ECE = %.4f | MCE = %.4f\n",
            ece_misc$ECE, ece_misc$MCE))
```

### All Metrics Summary Table

```{r}
#| label: tbl-all-metrics
#| tbl-cap: "Summary of all calibration metrics for well-calibrated and miscalibrated models. Consistent with Table 2 in Huang et al. (2020)."

# AUROC
auc_well <- auroc(p_well, y_test)
auc_misc <- auroc(p_misc, y_test)

# ICI via Loess
ici <- function(pred, truth) {
  df    <- tibble(pred, truth) %>% arrange(pred)
  loess_fit <- loess(truth ~ pred, data = df, span = 0.75)
  mean(abs(df$pred - predict(loess_fit, newdata = df)))
}

# Cox slope
cox_slope <- function(pred, truth) {
  df  <- tibble(y = truth, lp = qlogis(pmin(pmax(pred,1e-6),1-1e-6)))
  coef(glm(y ~ lp, data=df, family=binomial))[["lp"]]
}
cox_int <- function(pred, truth) {
  df  <- tibble(y = truth, lp = qlogis(pmin(pmax(pred,1e-6),1-1e-6)))
  coef(glm(y ~ offset(lp), data=df, family=binomial))[["(Intercept)"]]
}

tibble(
  Metric              = c("AUROC", "Brier Score", "H-L p-value",
                           "Cox Intercept (α)", "Cox Slope (β)",
                           "ECE", "MCE", "ICI"),
  `Well-calibrated`   = c(
    round(auc_well,  3),
    round(brier_well, 4),
    round(hl_well$p.value, 4),
    round(cox_int(p_well, y_test), 3),
    round(cox_slope(p_well, y_test), 3),
    round(ece_well$ECE, 4),
    round(ece_well$MCE, 4),
    round(ici(p_well, y_test), 4)
  ),
  `Miscalibrated`     = c(
    round(auc_misc,  3),
    round(brier_misc, 4),
    round(hl_misc$p.value, 4),
    round(cox_int(p_misc, y_test), 3),
    round(cox_slope(p_misc, y_test), 3),
    round(ece_misc$ECE, 4),
    round(ece_misc$MCE, 4),
    round(ici(p_misc, y_test), 4)
  ),
  `Ideal value`       = c("Higher is better", "0", "> 0.05",
                           "0", "1", "0", "0", "0")
) %>%
  knitr::kable(align = c("l","r","r","c"))
```

**Key observations from the table:**

- AUROC is nearly identical for both models — it completely misses the miscalibration.
- H-L p-value, Cox intercept, ECE, MCE, and ICI all correctly flag the miscalibrated model.
- Cox slope < 1 confirms the miscalibrated model overestimates high-risk patients and underestimates low-risk patients.

---


## Connection to Final Project (Hospital Readmission)

In the hospital readmission project, you will:

1. **Train on source domain** (`insurer == "Private"`) — model is calibrated for this population.

2. **Apply to target domain** (`insurer == "Medicare"`) — calibration may degrade due to:
    - Different baseline readmission rates (label shift).
    - Different patient mix and feature distributions (covariate shift).
    - Different relationships between predictors and readmission (concept shift).

3. **Measure degradation** using the full suite of metrics above on `target_test`.

4. **Recalibrate** using `target_cal` and re-evaluate on `target_test`.

The question your analysis will answer: *Which calibration metrics best detected the degradation, and which recalibration method most effectively corrected it?*

---

## What Comes Next

In the following sections we will cover:

- **Types of domain shift**: covariate shift, label shift, concept shift, and how to detect them.
- **Recalibration methods**: intercept-only recalibration, logistic recalibration (Platt scaling), isotonic regression.
- **Uncertainty quantification**: bootstrap confidence intervals for AUROC and Brier score.
- **Case study**: full before-vs-after analysis using the readmission dataset.
