---
title: "High-Dimensional Inference (Debiased / Desparsified Lasso)"
subtitle: "Biostat 274"
author: "Dr. Jin Zhou @ UCLA"
date: today
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    warning: FALSE
    cache: false
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
html-math-method: mathml
include-in-header:
  text: |
    <style>
      #title-slide .title {
        font-size: 4em;
        color: ;
      }
    </style>
execute:
  keep-md: false # Deletes intermediate markdown
  cache: false   # Disable caching if not needed
---

# Introduction

## Learning goals

- Why **lasso is great for prediction** but *not directly usable for inference*
- The **semi-low-dimensional** viewpoint: *parameter of interest + nuisance*
- Understand **orthogonalization / low-dimensional projection** (LDPE)
- Construct the **debiased / desparsified lasso** and interpret each term
- Build **Wald-style CIs and p-values** (plus multiple testing control)
- Run a **small simulation** to check:
  - coverage of 95% CIs
  - type I error / FDR behavior



---

## Why do we need new tools?

**Lasso achieves**

- shrinkage estimation
- variable selection
- strong prediction performance

**But lasso does NOT directly provide**

- valid p-values / confidence intervals for coefficients
- asymptotic normality for coefficients
- simple bootstrap-based uncertainty (exact zeros break smoothness)



---

## What is "high-dimensional inference" trying to achieve?

When $p > n$, goals include:

- **Unbiased, non-sparse** (or "less biased") estimator for target coefficients
- **Asymptotic normality** for low-dimensional functionals (e.g., a single $\beta_j$)
- **Accurate inference** (CIs, tests) for a small number of parameters



---

## Big picture: 5 families of high-dimensional inference

1) **Debiased / desparsified lasso** 
2) **Low-dimensional projection / orthogonal score** approaches  
3) **Sample splitting** / cross-fitting (easy, but lower power)  
4) **Selective inference** (condition on selection event)  
5) **Knockoffs** (FDR-controlled variable discovery)



---



## Semi-low-dimensional (LD) decomposition idea

A useful framing is

$$
\text{HD model} = \text{LD component} + \text{HD nuisance}.
$$

Example in linear regression $Y=X\beta+\varepsilon$:

- **LD (interest):** $X_{j_0}\beta_{j_0}$ (a "treatment effect" coefficient)
- **HD (nuisance):** $X_{-j_0}\beta_{-j_0}$



---

# Low-dimensional projection in the classical case

## Partialling-out when p is small

For coefficient $\beta_j$, the OLS estimator can be written using a residualized predictor:

$$
\hat\beta^{(lse)}_j
= \frac{(X_j^\perp)^\top Y}{(X_j^\perp)^\top X_j},
$$

where $X_j^\perp$ is the projection of $X_j$ onto the orthogonal complement of span $(X_{-j})$.



---

## What breaks when p > n?

When $p > n$, the column space of $X_{-j}$ fills $\mathbb{R}^n$, so

- the exact orthogonal residual $X_j^\perp$ can become **0**
- OLS becomes undefined / non-unique

Idea: replace "exact orthogonality" with "approximate orthogonality".



---

## Low-dimensional projection estimator (LDPE)

Pick a vector $z_j$ depending only on X, and take

$$
\hat\beta_j
=
\hat\beta^{(init)}_j
+
\frac{z_j^\top (Y - X\hat\beta^{(init)})}{z_j^\top X_j}.
$$

Interpretation:
- start from a sparse initial estimator $\hat\beta^{(init)}$ (often lasso)
- do a **one-step bias correction** using a linear projection of the residual



---

## Noise–bias decomposition (why $z_j$ matters)

A generic decomposition looks like

$$
\hat\beta_j-\beta_j
=
\underbrace{z_j^\top \varepsilon}_{\text{noise}}
+
\underbrace{\sum_{k\neq j} \frac{z_j^\top X_k}{z_j^\top X_j}\,(\beta_k-\hat\beta^{(init)}_k)}_{\text{bias from nuisance}}.
$$

So we must control:

- variance of $z_j^\top \varepsilon$
- correlations $z_j^\top X_k$ with nuisance features *and*
  the estimation error of $\hat\beta^{(init)}$


---

# Debiased / desparsified lasso

## The estimator (vector form)

Let $\hat\beta_{\lambda}$ be the lasso estimate, and $\hat\Sigma = X^\top X/n$.

Define

$$
\tilde\beta
=
\hat\beta_{\lambda}
+
\hat\Theta\,\frac{X^\top (Y - X\hat\beta_{\lambda})}{n},
$$

where $\hat\Theta$ is an estimate of $\Sigma^{-1}$ (precision matrix).

---

## A derivation sketch from the lasso KKT condition

In linear regression $Y=X\beta_0+\varepsilon$, lasso KKT implies

$$
\hat\Sigma(\hat\beta_\lambda-\beta_0) + \lambda \hat\tau
=
\frac{X^\top \varepsilon}{n}.
$$

If $\hat\Theta$ were exactly $\hat\Sigma^{-1}$, we’d get an (approximate) linear expansion:

$$
\tilde{\beta} - \beta_0 \approx \hat\Theta\,\frac{X^\top \varepsilon}{n}.
$$

The remainder term is controlled if $(\hat\Theta\hat\Sigma-I)$ is small enough and $\hat\beta_\lambda$ is sparse enough.



---

## How to estimate $\Theta \approx \Sigma^{-1}$

A standard approach: **nodewise lasso**.

For each j:

1) regress $X_j$ on $X_{-j}$ via lasso
2) use residuals to form row j of $\hat\Theta$

Intuition:

- in Gaussian designs, regression coefficients encode entries of the precision matrix
- sparsity of $\Theta$ means each feature depends on only a few others



---

## Asymptotic normality and CIs

Under regularity + sparsity conditions,

$$
\sqrt{n}(\tilde\beta-\beta_0)
=
W + o_p(1),
\quad
W\mid X \sim N_p\bigl(0,\ \sigma^2 \hat\Theta\,\hat\Sigma\,\hat\Theta^\top \bigr).
$$

So an approximate $100(1−\alpha)\%$ CI for $β_j$ is

$$
\tilde\beta_j \pm z_{1-\alpha/2}\ \hat\sigma\sqrt{\bigl(\hat\Theta\hat\Sigma\hat\Theta^\top\bigr)_{jj}/n}.
$$

---

## Alternative debiasing: Javanmard–Montanari

Rewrite debiased estimator as

$$
\tilde\beta
=
\hat\beta_\lambda
+
M\,\frac{X^\top(Y-X\hat\beta_\lambda)}{n}.
$$

Here $M$ does **not** have to be a perfect estimator of $\Sigma^{-1}$.

Instead, pick $M$ to balance:

- small bias: $\|M\hat\Sigma - I\|_\infty$ small
- small variance: diagonal entries of $M\hat\Sigma M^\top$ not too large


---

## Multiple testing

Once we have approximate Z-scores / p-values $\{p_j\}_{j=1}^p$, we can apply standard adjustments:

- **Bonferroni** for FWER control (very conservative)
- **Benjamini–Hochberg** for FDR control (often preferred in genomics)



---

# Extensions (optional)

## GLM and decorrelated score

The same ideas extend beyond Gaussian linear models.

One approach: **desparsified lasso for GLMs**

$$
\tilde\beta
=
\hat\beta_\lambda
+
\hat\Theta\ \frac{X^\top\{Y - b'(X\hat\beta_\lambda)\}}{n}.
$$

Another approach: **decorrelated score** functions, which explicitly build an orthogonal score for the low-dimensional parameter.



---

# Simulation

## Simulation design (matches common benchmarks)

We’ll simulate:

$$
Y = X\beta + \varepsilon,\quad
X \sim N(0,\Sigma),\quad
\varepsilon\sim N(0,\sigma^2 I).
$$

Two correlation structures:

- Toeplitz / AR(1): $\Sigma_{jk}=\rho^{|j-k|}$
- Equicorrelated: $\Sigma_{jk}=\rho$ for $j\neq k$

Targets:

- CI coverage for null and non-null coefficients
- type I error / FDR after BH


---

## R implementation plan

1) Generate $(X,y)$ with sparse $\beta$  
2) Fit **lasso** → $\hat\beta_\lambda$  
3) Fit **nodewise lasso** → $\hat\Theta$  
4) Compute **debiased** $\tilde\beta$ 
5) Compute SEs, CIs, p-values  
6) Repeat many times → empirical coverage / FDR



---

## ---- 0) Packages ----
```{r}
pkgs <- c("glmnet", "MASS", "Matrix")
for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) {
    install.packages(p, repos = "https://cloud.r-project.org")
  }
}
library(glmnet)
library(MASS)
library(Matrix)

set.seed(1)
```



---

## Helpers: covariance + sparse truth 
```{r}
toeplitz_cov <- function(p, rho = 0.7) {
  outer(1:p, 1:p, function(i, j) rho^abs(i - j))
}

equicorr_cov <- function(p, rho = 0.5) {
  Sigma <- matrix(rho, p, p)
  diag(Sigma) <- 1
  Sigma
}

make_beta <- function(p, s0 = 10, b = 1.0) {
  beta <- rep(0, p)
  beta[1:s0] <- b
  beta
}

standardize_X <- function(X) {
  Xs <- scale(X, center = TRUE, scale = TRUE)
  # rescale columns to ||X_j||_2 = sqrt(n)
  n <- nrow(Xs)
  Xs <- Xs * sqrt(n) / sqrt(colSums(Xs^2))
  Xs
}
```



---

## Fit initial lasso (CV) 
```{r}
fit_lasso_cv <- function(X, y) {
  # assume X,y centered; no intercept; no standardize (we already did)
  cv <- cv.glmnet(X, y, alpha = 1, intercept = FALSE, standardize = FALSE)
  beta <- as.vector(coef(cv, s = "lambda.min"))[-1]
  list(beta = beta, lambda = cv$lambda.min, cv = cv)
}
```



---

## Nodewise lasso to estimate Theta 
```{r}
nodewise_theta <- function(X, lambda_node = NULL) {
  # returns p x p matrix Theta_hat (dense in general)
  n <- nrow(X); p <- ncol(X)
  if (is.null(lambda_node)) {
    lambda_node <- 1.1 * sqrt(2 * log(p) / n)
  }

  Theta <- matrix(0, p, p)
  for (j in 1:p) {
    Xj <- X[, j]
    Xmj <- X[, -j, drop = FALSE]

    fit <- glmnet(Xmj, Xj, alpha = 1, lambda = lambda_node,
                  intercept = FALSE, standardize = FALSE)
    gamma <- as.vector(coef(fit))[-1]  # length p-1

    rj <- as.vector(Xj - Xmj %*% gamma)

    # tau^2 estimator commonly used in nodewise lasso:
    tau2 <- mean(rj^2) + lambda_node * sum(abs(gamma))

    Theta[j, j] <- 1 / tau2
    Theta[j, -j] <- -gamma / tau2
  }
  Theta
}
```



---

## Debiased estimator + SEs 
```{r}
debiased_lasso <- function(X, y, sigma = NULL, lambda_node = NULL) {
  n <- nrow(X); p <- ncol(X)

  # initial lasso
  fit0 <- fit_lasso_cv(X, y)
  beta_lasso <- fit0$beta

  # variance estimate (for simulation we often use the true sigma)
  if (is.null(sigma)) {
    resid <- as.vector(y - X %*% beta_lasso)
    sigma <- sqrt(mean(resid^2))
  }

  # Theta estimate
  Theta <- nodewise_theta(X, lambda_node = lambda_node)

  # debias
  resid <- as.vector(y - X %*% beta_lasso)
  beta_db <- beta_lasso + as.vector(Theta %*% (crossprod(X, resid) / n))

  # SEs from sigma^2 * (Theta Sigma Theta^T) / n
  Sigma_hat <- crossprod(X) / n
  V <- Theta %*% Sigma_hat %*% t(Theta)
  se <- sigma * sqrt(pmax(diag(V), 0) / n)

  list(beta_lasso = beta_lasso, beta_db = beta_db, se = se,
       sigma = sigma, Theta = Theta, Sigma_hat = Sigma_hat,
       lambda_lasso = fit0$lambda)
}
```



---

## Single run demo: CI plot for selected coefficients 
```{r}
one_run_demo <- function(n = 120, p = 300, s0 = 10, b = 1, sigma = 1,
                         Sigma_type = c("toeplitz", "equicorr"), rho = 0.7) {

  Sigma_type <- match.arg(Sigma_type)
  Sigma <- if (Sigma_type == "toeplitz") toeplitz_cov(p, rho) else equicorr_cov(p, rho)

  beta0 <- make_beta(p, s0, b)

  X <- MASS::mvrnorm(n, mu = rep(0, p), Sigma = Sigma)
  X <- standardize_X(X)

  eps <- rnorm(n, 0, sigma)
  y <- as.vector(X %*% beta0 + eps)
  y <- scale(y, center = TRUE, scale = FALSE)

  out <- debiased_lasso(X, y, sigma = sigma)

  # show a handful of coefficients (signals + nulls)
  idx <- c(1:5, (s0 + 1):(s0 + 5))
  ci_low <- out$beta_db[idx] - 1.96 * out$se[idx]
  ci_high <- out$beta_db[idx] + 1.96 * out$se[idx]

  data.frame(
    j = idx,
    beta_true = beta0[idx],
    beta_db = out$beta_db[idx],
    se = out$se[idx],
    ci_low = ci_low,
    ci_high = ci_high,
    covered = (ci_low <= beta0[idx]) & (beta0[idx] <= ci_high)
  )
}

demo_df <- one_run_demo(n = 120, p = 300, s0 = 10, b = 1, sigma = 1,
                        Sigma_type = "toeplitz", rho = 0.7)
demo_df
```



---

## Repeated simulation: empirical coverage + BH-FDR 
```{r, eval = F}
sim_replicates <- function(B = 30, n = 120, p = 300, s0 = 10, b = 1, sigma = 1,
                           Sigma_type = c("toeplitz", "equicorr"), rho = 0.7,
                           alpha = 0.05) {

  Sigma_type <- match.arg(Sigma_type)
  Sigma <- if (Sigma_type == "toeplitz") toeplitz_cov(p, rho) else equicorr_cov(p, rho)
  beta0 <- make_beta(p, s0, b)
  S <- which(beta0 != 0)

  cover_S <- numeric(B)
  cover_null <- numeric(B)
  fdr <- numeric(B)
  power <- numeric(B)

  for (bb in 1:B) {
    X <- MASS::mvrnorm(n, mu = rep(0, p), Sigma = Sigma)
    X <- standardize_X(X)

    y <- as.vector(X %*% beta0 + rnorm(n, 0, sigma))
    y <- scale(y, center = TRUE, scale = FALSE)

    out <- debiased_lasso(X, y, sigma = sigma)

    ci_low <- out$beta_db - qnorm(1 - alpha/2) * out$se
    ci_high <- out$beta_db + qnorm(1 - alpha/2) * out$se

    cover_S[bb] <- mean(ci_low[S] <= beta0[S] & beta0[S] <= ci_high[S])
    cover_null[bb] <- mean(ci_low[-S] <= 0 & 0 <= ci_high[-S])

    z <- out$beta_db / out$se
    pval <- 2 * (1 - pnorm(abs(z)))

    # BH discoveries
    qval <- p.adjust(pval, method = "BH")
    disc <- which(qval <= alpha)

    if (length(disc) == 0) {
      fdr[bb] <- 0
      power[bb] <- 0
    } else {
      fdr[bb] <- mean(!(disc %in% S))
      power[bb] <- mean(S %in% disc)
    }
  }

  data.frame(
    Sigma_type = Sigma_type,
    n = n, p = p, s0 = s0, b = b, sigma = sigma, rho = rho, B = B,
    cover_S_mean = mean(cover_S),
    cover_null_mean = mean(cover_null),
    fdr_mean = mean(fdr),
    power_mean = mean(power)
  )
}

res_toeplitz <- sim_replicates(B = 20, Sigma_type = "toeplitz", rho = 0.7)
res_equicorr <- sim_replicates(B = 20, Sigma_type = "equicorr", rho = 0.5)

rbind(res_toeplitz, res_equicorr)
```



---

## Practical cautions (what to tell students)

- **Variance estimation $\sigma$** matters (scaled lasso, refit, or other estimators)
- **Strong correlations** can hurt: Θ estimation becomes unstable
- Many results assume sparsity in **$\beta$** and/or in rows of **$\Theta$**
- For genomics: always report **BH q-values** (or other error-rate control), not raw p-values
- “Inference after feature screening” needs care (screening changes the target)



---

## Summary

- High-dimensional inference focuses on **low-dimensional targets** with high-dimensional nuisance
- Orthogonalization / projection yields **debiased** estimators with asymptotic normality
- The debiased lasso is a practical one-step correction:

$$
\tilde\beta
=
\hat\beta_\lambda + \hat\Theta X^\top (Y - X\hat\beta_\lambda)/n
$$

- Simulation is a sanity check: **coverage** and **FDR/power**



---

# Suggested readings

- Fan et al. 2020, Statistical Foundations of Data Science (Ch. 7, esp. 7.1, 7.2), for a comprehensive review
- van de Geer et al. (2014), Javanmard & Montanari (2014) for foundational theory
- Dezeure et al. (2015) for the R package `hdi` and practical comparisons


